---
title: 'Data import, quality control and normalization for the Cuomo dataset'
author: "Koen Van den Berge and Jeroen Gilis"
date: "20/11/2021"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Introduction

We here make use of the publication of Anna Cuomo et al.
(last author Oliver Stegle), which we will refer to as the `iPSC dataset`. The 
paper that describes this dataset can be found using this 
[link](https://www.nature.com/articles/s41467-020-14457-z).

In the experiment, the authors harvested induced pluripotent stem cells (iPSCs)
from 125 healthy human donors. These cells were used to study the endoderm 
differentiation process. In this process, iPSCs differentiate to endoderm cells,
a process which takes approximately three days. As such, the authors 
cultered the iPSCs cell lines and allowed for differentiation for three days. 
During the experiment, cells were harvested at four different time points: 
day0 (directly at to incubation), day1, day2 and day3. Knowing the process of 
endoderm differentiation, these time points should correspond with different 
cell types: day0 are (undifferentiated) iPSCs, day1 are mesendoderm cells, day2
are "intermediate" cells and day3 are fully differentiated endoderm cells.

This dataset was generated using the **SMART-Seq2** scRNA-seq protocol.

The final goal of the experiment was to characterize population variation in the
process of endoderm differentiation.

# Download data

For this lab session, we will work with a subset of the data, i.e., the data
for the first (alphabetically) 15 patients in the experiment. These can be
downloaded through the *belnet filesender* link provided through email,
https://filesender.belnet.be/?s=download&token=eb8136df-67d3-4869-b2a9-f65767054e81.

The data original (125 patient) could be downloaded from 
[Zenodo](https://zenodo.org/record/3625024#.YWfahtlBxB1). At the bottom of this
web-page, we can download the files `raw_counts.csv.zip` and 
`cell_metadata_cols.tsv` and store these files locally. We do not recommend 
doing this during the lab session, to avoid overloading the system.

# Import data

First we read in the count matrix:

```{r, message=FALSE, warning=FALSE}
sce <- readRDS("/Users/jg/Desktop/sce_15_cuomo.rds")
```

```{r}
sce
```

# Explore metadata

Exploration of the metadata is essential to get a better idea of what the
experiment was about and how it was organized.

```{r}
colData(sce)[1:5,1:10]
colnames(colData(sce))
```

As stated in the paper, cells were sampled on 4 time points. Each of these 
time points is expected to correspond with different cell types (day0 = iPSC,
day1 = mesendoderm, day2 = intermediate and day3 = endoderm).

```{r}
table(colData(sce)$day)
```

As stated in the paper, cells were harvested from 125 patients. Here, we are
working on a subset with 15 patients. The number of cells harvested per patient 
(over all time points) ranges from 31 to 637.

```{r}
length(table(colData(sce)$donor)) # number of donors
range(table(colData(sce)$donor)) # cells per donor
```

Below, we look how many cells are harvest per patent and per time point.

```{r}
table(colData(sce)$donor,colData(sce)$day)
```

We see that for many patients the data is complete, i.e. cells were sampled
on all time points.

Practically, the cells were prepared in 28 batches. Since we here only look
at a subset of the data, we see that only 14 of these batches are represented 
here.

```{r}
length(table(colData(sce)$experiment))
table(colData(sce)$experiment, colData(sce)$day)
```

# Obtaining and including rowData

The `rowData` slot of a `SingleCellExperiment` object allows for storing 
information on the features, i.e. the genes, in a dataset. In our object,
the `rowData` slot currently contains the following:

```{r}
head(rowData(sce))
```

To improve our gene-level information, we may:

1. Split `V1` into two columns, one with the ENSEMBL ID and the other with 
the gene symbol.

2. Display which chromosome the gene is located

Many more options are possible, but are not necessary for us right now.

```{r}
rowData(sce) <- data.frame(Ensembl = gsub("_.*", "", rowData(sce)$V1),
                           Symbol = gsub("^[^_]*_", "", rowData(sce)$V1))
head(rowData(sce))
```


```{r, message=FALSE, warning=FALSE}
library("biomaRt")
ensembl75 <- useEnsembl(biomart = 'genes', 
                        dataset = 'hsapiens_gene_ensembl',
                        version = 75)

GeneInfo <- getBM(attributes = c("ensembl_gene_id", # To match with rownames SCE
                                 "chromosome_name"), # Info on chromose
                  mart = ensembl75)
GeneInfo <- GeneInfo[match(rowData(sce)$Ensembl, GeneInfo$ensembl_gene_id),]

rowData(sce) <- cbind(rowData(sce), GeneInfo)
head(rowData(sce))
all(rowData(sce)$Ensembl == rowData(sce)$ensembl_gene_id) 
# identical, as desired, so we could optionally remove one of the two
```

# Filtering non-informative genes

Let us first try the very simple and very lenient filtering criterion that we
adopted for the Macosko dataset.

```{r}
keep <- rowSums(assays(sce)$counts > 0) > 10
table(keep)

sce <- sce[keep,]
```

We see that this filtering strategy does not remove any genes for this dataset.
In general, datasets from plate-based scRNA-seq dataset have a far higher
sequencing depth than data from droplet-based protocols. As requiring a minimum
expression of 1 count in at least 10 cells is a very lenient criterion if we 
consider that we have 36.000 cells, we should consider adopting a more stringent
filtering criterium, like the `filterByExpr` from `edgeR`:

```{r, message=FALSE, warning=FALSE}
library(edgeR)

table(colData(sce)$day)

keep2 <- edgeR::filterByExpr(y=sce,
                             group = colData(sce)$day,
                             min.count = 5,
                             min.prop = 0.4)
table(keep2)

sce <- sce[keep2,]
```

# Quality control

## Calculate QC variables

```{r}
library(scater)

# check ERCC spike-in transcripts
sum(grepl("^ERCC-", rowData(sce)$Symbol)) # no spike-in transcripts available

is.mito <- grepl("^MT", rowData(sce)$chromosome_name)
sum(is.mito) # 13 mitochondrial genes

df <- perCellQCMetrics(sce, subsets=list(Mito=is.mito))
head(df)

## add the QC variables to sce object
colData(sce) <- cbind(colData(sce), df)
```

## Exploratory data analysis

In the figure below, we see that several cells have a very low number of 
expressed genes, and where most of the molecules are derived from 
mitochondrial genes. This indicates likely damaged cells, presumably because 
of loss of cytoplasmic RNA from perforated cells, so we should remove these for 
the downstream analysis.

```{r}
# Number of genes vs library size
plotColData(sce, x = "sum", y="detected", colour_by="day") 

# Mitochondrial genes
plotColData(sce, x = "detected", y="subsets_Mito_percent", colour_by="day")
```

## QC using adaptive thresholds

Below, we remove cells that are outlying with respect to

 1. A low sequencing depth (number of UMIs);
 2. A low number of genes detected;
 3. A high percentage of reads from mitochondrial genes.
 
We remove a total of $301$ cells, mainly due to low sequencing depth and
low number of genes detected.

```{r}
lowLib <- isOutlier(df$sum, type="lower", log=TRUE)
lowFeatures <- isOutlier(df$detected, type="lower", log=TRUE)
highMito <- isOutlier(df$subsets_Mito_percent, type="higher")

table(lowLib)
table(lowFeatures)
table(highMito)

discardCells <- (lowLib | lowFeatures | highMito)
table(discardCells)
colData(sce)$discardCells <- discardCells

# visualize cells to be removed
plotColData(sce, x = "detected", y="subsets_Mito_percent", colour_by = "discardCells")
plotColData(sce, x = "sum", y="detected", colour_by="discardCells")
```

```{r}
# visualize cells to be removed
plotColData(sce, x = "detected", y="subsets_Mito_percent", colour_by = "donor")
plotColData(sce, x = "sum", y="detected", colour_by="donor")
```

```{r}
# visualize cells to be removed
plotColData(sce, x = "detected", y="subsets_Mito_percent", colour_by = "experiment")
plotColData(sce, x = "sum", y="detected", colour_by="experiment")
```

```{r}
table(sce$donor, sce$discardCells)
table(sce$donor, sce$discardCells)/rowSums(table(sce$donor, sce$discardCells))
#fractions of removed cells per donor
```

Most removed cells (fraction) are from patients `dixh` and `babz`.

```{r}
table(sce$experiment, sce$discardCells)
table(sce$experiment, sce$donor)
```

Most removed cells (fraction) are from patients `dixh` and `babz`.
Most low library sizes seem to come from patient `dixh`; for patient `babz`
the effect is less pronounced.

```{r}
plotColData(sce[,sce$donor=="dixh"], x = "sum", y="detected")
plotColData(sce[,sce$donor=="babz"], x = "sum", y="detected")
```

As such, we are mainly removing cells from specific patients and the respective
batches in which they were sequenced. However, we want to be careful; we only
want to remove technical artefacts, while retaining as much of the biology as
possible. In our exploratory figure, we see that the cells we are removing based
on the number of genes detected, are quite far apart from the bulk of the data
cloud; as such, these cells are indeed suspicious. For the criterion of
library size, we see that the cells removed there are still strongly connected
to the data cloud. As such, we may want to relax the filtering criterion there a
little bit. When we think of how the adaptive threshold strategy works, we
may want to remove cells that are 4MADs away from the center, rather than
the default 3 MADs.

```{r}
# previously
lowLib <- isOutlier(df$sum, type="lower", log=TRUE)
table(lowLib)

# after seeing appropriate exploratory figure
lowLib <- isOutlier(df$sum, nmads=4, type="lower", log=TRUE)
table(lowLib)

discardCells <- (lowLib | lowFeatures | highMito)
table(discardCells)
colData(sce)$discardCells <- discardCells
```

Note that these steps are not exact; different analysts will come with different
filtering criteria for many of the steps. The key ideas are that
we let appropriate exploratory figures guide us to make reasonable choices;
i.e., we look at the data rather than blindly following a standardized pipeline
that may work well in many cases, but maybe not our particular dataset.

```{r}
# remove cells identified using adaptive thresholds
sce <- sce[, !colData(sce)$discardCells]
```

## Identifying and removing empty droplets

This does not make much sense for plate-based data!

```
library(DropletUtils)
bcrank <- barcodeRanks(counts(sce))

# Only showing unique points for plotting speed. Duplicated ranks are a 
# consequence of ties in the ranks, i.e., when cells have an equal library size.
sum(duplicated(bcrank$rank))
uniq <- !duplicated(bcrank$rank)

plot(bcrank$rank[uniq], bcrank$total[uniq], log="xy",
    xlab="Rank", ylab="Total UMI count", cex.lab=1.2)
abline(h=metadata(bcrank)$inflection, col="darkgreen", lty=2)
abline(h=metadata(bcrank)$knee, col="dodgerblue", lty=2)
abline(h=3e+5, col="orange", lty=2) # picked visually myself
legend("topright", 
       legend=c("Inflection", "Knee", "Empirical knee point"), 
       col=c("darkgreen", "dodgerblue", "orange"), 
       lty=2, 
       cex=1.2)

set.seed(100)
limit <- 3e+5   
all.out <- emptyDrops(counts(sce), lower=limit, test.ambient=TRUE)
# p-values for cells with total UMI count under the lower bound.
hist(all.out$PValue[all.out$Total <= limit & all.out$Total > 0],
    xlab="P-value", main="", col="grey80", breaks=20)

# but note that it would remove a very high number of cells
length(which(all.out$FDR <= 0.01)) # retained
```

## Identifying and removing doublets

We will use [scDblFinder](https://bioconductor.org/packages/3.14/bioc/html/scDblFinder.html) to detect doublet cells.

```{r, message=FALSE, warning=FALSE}
## perform doublet detection
library(scDblFinder)
```

```{r}
set.seed(7498864)
sce <- scDblFinder(sce, 
                   samples = as.factor(sce$experiment), # 14 batches
                   returnType="sce")
## visualize these scores
## explore doublet score wrt original day labels
boxplot(log1p(sce$scDblFinder.score) ~ factor(colData(sce)$day, exclude=NULL))

tab <- table(sce$scDblFinder.class, sce$day, 
      exclude=NULL)
tab
t(t(tab) / colSums(tab))

barplot(t(t(tab) / colSums(tab))[2,],
        xlab = "Cluster", ylab = "Fraction of doublets")
```

Only a very small fraction of the cells are removed in this step. Indeed, this
step is much more relevant for droplet-based protocols. For plate-based 
protocols like smartseq2, the experimental procedure usually takes care of
removing empty wells, and lasers are also used to detect wells with more than
1 cell.

```
# remove doublets -> I will here not remove them
sce <- sce[,!sce$scDblFinder.class == "doublet"]
```

# Normalization

For normalization, the size factors $s_i$ computed here are simply scaled 
library sizes:

\[ N_i = \sum_g Y_{gi} \]
\[ s_i = N_i / \bar{N}_i \]

```{r}
sce <- logNormCounts(sce)

# note we also returned log counts: see the additional logcounts assay.
sce

# you can extract size factors using
sf <- librarySizeFactors(sce)
mean(sf) # equal to 1 due to scaling.
plot(x= log(colSums(assays(sce)$counts)), 
     y=sf)
```

--- end lab session 1 ---

